---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---
## Reproducible Research Peer Assessment - 1


## Loading and preprocessing the data

```{r}
## Unzipping the activity Data file and then reading the .csv file into a Data Frame


zipFileName <- "./repdata_data_activity.zip"
unzip(zipFileName)
activityDF <- read.csv("./activity.csv")

# converting the date variable into date format 

activityDF$date <-as.Date(activityDF$date)

# summary(activityDF)
# str(activityDF)

# Check the number of rows in the activity data file

nrow(activityDF)
head(activityDF)

```

## What is mean total number of steps taken per day?

``` {r}

# For this step we need to ignore NA cases

activityCompCases <- activityDF[complete.cases(activityDF),]

print ("Original Rows:")
print (nrow(activityDF))

print ("After removing NAs:")
print(nrow(activityCompCases))

dateWiseStepsSum <- aggregate(activityCompCases$steps, list(activityCompCases$date), FUN = sum)
colnames(dateWiseStepsSum) <- c("Date", "TotSteps")

print("Date Wise Total Steps:")
print(dateWiseStepsSum)

## Plot of Total Number of Steps Taken Every Day

library(ggplot2)
stepsPlot <- ggplot(activityCompCases, aes(x = as.factor(date), y= steps)) 

stepsPlot + geom_bar (stat = "identity") + theme(axis.text.x = element_text(angle=-90)) + xlab ("Date") + ylab ("Total Steps")


## The histogram of total steps taken each day 

xMax <- max(dateWiseStepsSum$TotSteps) + 7000


hist(dateWiseStepsSum$TotSteps, breaks = 15, xlim = c(0, xMax), xlab = "Days::Total Steps", main = "Histogram of Days-Total Steps", col = "light blue")

## The mean and median of the total number of steps taken every day

#dateWiseStepsMean <- aggregate(activityCompCases$steps, list(activityCompCases$date), FUN=mean)
#print("Mean of Steps every day:")
#print(dateWiseStepsMean)

#dateWiseStepsMedian <- aggregate(activityCompCases$steps, list(activityCompCases$date), FUN=median)
#print("Median of Steps every day:")
#print(dateWiseStepsMedian)


print("Mean of the total number of steps taken per day:")
print(mean(dateWiseStepsSum$TotSteps))

print("Median of the total number of steps taken per day:")
print(median(dateWiseStepsSum$TotSteps))

```

## What is the average daily activity pattern?
 
```{r}

## Interval Wise Mean of Steps across all days

IntervalWiseStepsMean <- aggregate(activityCompCases$steps, list(activityCompCases$interval), FUN=mean)

colnames(IntervalWiseStepsMean) <- c("Interval", "AvgSteps")

ggplot(data = IntervalWiseStepsMean, aes(x=Interval, y=AvgSteps)) + geom_line() + geom_point() + xlab ("Interval") + ylab ("Average Steps")

## The 5-minute interval, that contains the maximum number of steps

print("The 5-minute interval, that contains the maximum number of steps is:")
maximumIndex <- which.max(IntervalWiseStepsMean$AvgSteps)
print(IntervalWiseStepsMean[maximumIndex,"Interval"])


```

## Imputing missing values

```{r}

## Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

print ("Number of missing values:")
print(length(which(is.na(activityDF$steps))))

## Lets use Average for a time interval to populate missing values in the original data

NewActivityDF <- activityDF

for (i in 1:nrow(NewActivityDF)) {
  if (is.na(NewActivityDF[i,"steps"])) {
    timeInt <- NewActivityDF[i, "interval"]
    AvgStepsOfInt <- IntervalWiseStepsMean[IntervalWiseStepsMean$Interval==timeInt, "AvgSteps"]
    NewActivityDF[[i,"steps"]] <- AvgStepsOfInt
  }
}

NewdateWiseStepsSum <- aggregate(NewActivityDF$steps, list(NewActivityDF$date), FUN = sum)
colnames(NewdateWiseStepsSum) <- c("Date", "TotSteps")

xMax <- max(NewdateWiseStepsSum$TotSteps) + 7000


hist(NewdateWiseStepsSum$TotSteps, breaks = 15, xlim = c(0, xMax), xlab = "Days::Total Steps", main = "New Histogram of Days-Total Steps", col = "blue")


print("New Mean of the total number of steps taken per day:")
print(mean(NewdateWiseStepsSum$TotSteps))

print("New Median of the total number of steps taken per day:")
print(median(NewdateWiseStepsSum$TotSteps))

## Marginal increase in the mean, median values
## Frequency counts for total steps increased in the histogram

```

### Marginal increase in the mean, median values
### Frequency counts for total steps increased in the histogram

## Are there differences in activity patterns between weekdays and weekends?

```{r}
WeekType <- data.frame()

for (i in 1:nrow(NewActivityDF)) {
  if ((weekdays(NewActivityDF [i, "date"]) == "Sunday") || (weekdays(NewActivityDF [i, "date"]) == "Saturday")) {
    WeekType[i,"WD"] <- "Weekend"
  } else {
    WeekType[i,"WD"] <- "WeekDay"
  }
}

cData <- cbind(NewActivityDF, WeekType)

sData <- aggregate(cData$steps, by = list(cData$WD, cData$interval), FUN = sum)

colnames(sData) <- c("WD", "Interval", "TotSteps")

ggplot(data=sData, aes(x=Interval, y=TotSteps)) + 
    geom_line() + xlab("Interval") +
    ylab("Average number of steps") + facet_grid(WD ~ .)


```

### See higher counts during weekdays compared to weekends
